---
title: "Dataset Overview"
page-layout: full
---

```{r}
library(xml2)
library(jsonlite)

eml_lines <- readLines("data/derived/dwc/eml-waterbird_counts_sabaki_mida.xml", warn = FALSE)
eml_lines <- gsub(
  "&(?!(amp;|lt;|gt;|quot;|apos;|#))",
  "&amp;",
  eml_lines,
  perl = TRUE
)
eml <- read_xml(paste(eml_lines, collapse = "\n"))
title_text <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='dataset']/*[local-name()='title']"
))
creator_nodes <- xml_find_all(
  eml,
  "//*[local-name()='dataset']/*[local-name()='creator']"
)
creators <- vapply(
  creator_nodes,
  function(node) {
    given <- xml_text(xml_find_first(node, ".//*[local-name()='givenName']"))
    sur <- xml_text(xml_find_first(node, ".//*[local-name()='surName']"))
    paste(given, sur)
  },
  character(1)
)

abstract_paras <- xml_find_all(
  eml,
  "//*[local-name()='dataset']/*[local-name()='abstract']/*[local-name()='para']"
) |>
  xml_text()

pub_date <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='dataset']/*[local-name()='pubDate']"
))
doi_url <- xml_find_all(
  eml,
  "//*[local-name()='dataset']/*[local-name()='alternateIdentifier']"
) |>
  xml_text() |>
  (\(x) x[grepl("^https://doi.org/", x)][1])()

begin_date <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='beginDate']/*[local-name()='calendarDate']"
))
end_date <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='endDate']/*[local-name()='calendarDate']"
))
west <- xml_text(xml_find_first(eml, "//*[local-name()='westBoundingCoordinate']"))
east <- xml_text(xml_find_first(eml, "//*[local-name()='eastBoundingCoordinate']"))
north <- xml_text(xml_find_first(eml, "//*[local-name()='northBoundingCoordinate']"))
south <- xml_text(xml_find_first(eml, "//*[local-name()='southBoundingCoordinate']"))

latest_events <- list.files(
  "data/derived/dwc",
  pattern = "^events_\\d{4}-\\d{2}-\\d{2}\\.csv$",
  full.names = FALSE
)
latest_events <- latest_events[order(latest_events, decreasing = TRUE)][1]

latest_occurrences <- list.files(
  "data/derived/dwc",
  pattern = "^occurrences_\\d{4}-\\d{2}-\\d{2}\\.csv$",
  full.names = FALSE
)
latest_occurrences <- latest_occurrences[order(latest_occurrences, decreasing = TRUE)][1]

summary_cards <- jsonlite::fromJSON("data/derived/dashboard/summary_cards.json")
```

::: {.hero}
# `r title_text`

Long-term waterbird monitoring data from Sabaki River Mouth and Mida Creek, maintained by A Rocha Kenya. This page is the entry point; the numbered scripts contain the executable data checks and export steps.

```{r, results="asis"}
cat(
  paste0(
    '<div class="hero-links">',
    '<a class="hero-link" href="', doi_url, '">GBIF DOI</a>',
    '<a class="hero-link" href="dashboard/">Interactive dashboard</a>',
    '<a class="hero-link" href="script/01_data_checks.html">01 Data checks</a>',
    '<a class="hero-link" href="script/02_export_outputs.html">02 Export outputs</a>',
    '</div>'
  )
)
```
:::

::: {.overview-grid}
::: {.overview-card}
**Coverage**

`r begin_date` to `r end_date`
:::

::: {.overview-card}
**Surveys**

`r summary_cards$events`
:::

::: {.overview-card}
**Taxa**

`r summary_cards$taxa`
:::

::: {.overview-card}
**Individuals**

`r scales::comma(summary_cards$individuals)`
:::
:::

## Use the Dataset

::: {.link-grid}
::: {.link-card}
### Explore

[Open the dashboard](dashboard/) for filtering by site, year, month, and taxon.
:::

::: {.link-card}
### Verify

[Review the data checks script](script/01_data_checks.html) before using the export tables.
:::

::: {.link-card}
### Reproduce

[Read the export script](script/02_export_outputs.html) to see how Darwin Core files, dashboard JSON, and EML fields are generated.
:::
:::

## Current Export

The current metadata record was refreshed on `r pub_date`. The latest Darwin Core files are:

```{r, results="asis"}
cat(
  paste(
    c(
      paste0("- [Events table](data/derived/dwc/", latest_events, ")"),
      paste0("- [Occurrences table](data/derived/dwc/", latest_occurrences, ")"),
      "- [EML metadata](data/derived/dwc/eml-waterbird_counts_sabaki_mida.xml)",
      "- [Field notebook instructions](assets/Notebook_Instruction.pdf)"
    ),
    collapse = "\n"
  )
)
```

## Metadata Source

The homepage title, creators, abstract summary, DOI, publication date, temporal coverage, and bounding box are read from the generated EML XML file in `data/derived/dwc/`.

**Creators:** `r paste(creators, collapse = ", ")`

**Bounding box:** `r south` to `r north` latitude; `r west` to `r east` longitude.

```{r, results="asis"}
cat(abstract_paras[1])
```
