---
title: "Dataset Overview"
page-layout: full
---

```{r}
library(xml2)
library(jsonlite)

eml_lines <- readLines(
  "data/derived/dwc/eml-waterbird_counts_sabaki_mida.xml",
  warn = FALSE
)
eml_lines <- gsub(
  "&(?!(amp;|lt;|gt;|quot;|apos;|#))",
  "&amp;",
  eml_lines,
  perl = TRUE
)
eml <- read_xml(paste(eml_lines, collapse = "\n"))
abstract_paras <- xml_find_all(
  eml,
  "//*[local-name()='dataset']/*[local-name()='abstract']/*[local-name()='para']"
) |>
  xml_text()

doi_url <- xml_find_all(
  eml,
  "//*[local-name()='dataset']/*[local-name()='alternateIdentifier']"
) |>
  xml_text() |>
  (\(x) x[grepl("^https://doi.org/", x)][1])()

begin_date <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='beginDate']/*[local-name()='calendarDate']"
))
end_date <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='endDate']/*[local-name()='calendarDate']"
))
citation_text <- xml_text(xml_find_first(
  eml,
  "//*[local-name()='additionalMetadata']//*[local-name()='gbif']/*[local-name()='citation']"
))
coverage_years <- paste(
  format(as.Date(begin_date), "%Y"),
  format(as.Date(end_date), "%Y"),
  sep = " to "
)

latest_events <- list.files(
  "data/derived/dwc",
  pattern = "^events_\\d{4}-\\d{2}-\\d{2}\\.csv$",
  full.names = FALSE
)
latest_events <- latest_events[order(latest_events, decreasing = TRUE)][1]

latest_occurrences <- list.files(
  "data/derived/dwc",
  pattern = "^occurrences_\\d{4}-\\d{2}-\\d{2}\\.csv$",
  full.names = FALSE
)
latest_occurrences <- latest_occurrences[order(
  latest_occurrences,
  decreasing = TRUE
)][1]

summary_cards <- jsonlite::fromJSON("data/derived/dashboard/summary_cards.json")
```

::: {.hero}
`r abstract_paras[1]`

```{r, results="asis"}
cat(
  paste0(
    '<div class="hero-links">',
    '<a class="hero-link" href="',
    doi_url,
    '">View on GBIF</a>',
    '</div>'
  )
)
```
:::

::: {.overview-grid}
::: {.overview-card}
**Coverage**

`r coverage_years`
:::

::: {.overview-card}
**Surveys**

`r summary_cards$events`
:::

::: {.overview-card}
**Taxa**

`r summary_cards$taxa`
:::

::: {.overview-card}
**Individuals**

`r scales::comma(summary_cards$individuals)`
:::
:::

## How is the data collected?

1. [Fieldnote](assets/Notebook_Instruction.pdf)
   The observers record species, counts, timing, and field conditions in the notebook during each survey.
2. [Entering in Excel](https://github.com/A-Rocha-Kenya/Waterbird-counts-Sabaki-Mida/blob/master/data/raw/water_bird_count_data.xlsx)
   The field notes are transcribed into the master spreadsheet that stores the long-term dataset.
3. [Data validation check](script/01_data_checks.html)
   The Quarto checks look for missing values, inconsistent event metadata, lookup mismatches, and duplicates.
4. [Export to Darwin Core](script/02_export_outputs.html)
   The cleaned spreadsheet is converted into publication-ready Darwin Core, EML, and dashboard files.

## How to use the data?

- [Use the Excel file](https://github.com/A-Rocha-Kenya/Waterbird-counts-Sabaki-Mida/blob/master/data/raw/water_bird_count_data.xlsx)
  This is the most direct version of the source data for inspection and manual reuse.
- [Use the events CSV](data/derived/dwc/`r latest_events`)
  This file contains one row per survey event with dates, sites, timing, and event-level metadata.
- [Use the occurrence CSV](data/derived/dwc/`r latest_occurrences`)
  This file contains one row per taxon record linked to survey events through `eventID`.
- [Use the Darwin Core Archive on GBIF](`r doi_url`)
  This is the published version of the dataset with metadata and standard biodiversity packaging.

## Citation

Please cite the dataset as:

> `r citation_text`
