---
title: "01 Data checks"
author: "Raphaël Nussbaumer"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
source(here::here("R", "load_data.R"))
source(here::here("R", "load_helper.R"))

load_helper()
data_objects <- load_data()
list2env(data_objects, envir = knitr::knit_global())
```

## Introduction

This document assembles the core quality checks used before publishing the waterbird count dataset for Sabaki River Mouth and Mida Creek.

## Basic statistics

```{r basic-statistics}
summary_table <- bind_rows(
  counts |> mutate(site_original = site),
  counts |> mutate(site_original = site, site = "Combined")
) |>
  group_by(site) |>
  summarise(
    n_survey = n_distinct(date, site_original),
    n_taxon = n_distinct(common_name),
    n_observation = n(),
    n_individual = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

show_table(
  summary_table,
  caption = htmltools::tags$caption(
    "Summary of surveys, taxa, observations, and individuals."
  )
)
```

## Quality control

### Assessment 1: One metadata record per event

```{r qc-event-metadata}
event_metadata_check <- counts |>
  group_by(date, site) |>
  summarise(
    n_start_time = n_distinct(start_time),
    n_end_time = n_distinct(end_time),
    n_coverage = n_distinct(coverage),
    n_method = n_distinct(method),
    n_water = n_distinct(water),
    n_tidal = n_distinct(tidal),
    n_weather = n_distinct(weather),
    n_disturbed = n_distinct(disturbed),
    n_participants = n_distinct(participants),
    .groups = "drop"
  ) |>
  filter(if_any(starts_with("n_"), ~ .x > 1))

show_check(event_metadata_check)
```

### Assessment 2: Missing species name or count

```{r qc-missing-values}
missing_count_check <- counts |> filter(is.na(count))
missing_name_check <- counts |> filter(is.na(common_name))

show_check(missing_count_check, "No missing counts found.")
cat("\n\n")
show_check(missing_name_check, "No missing species names found.")

stop_if_issues(
  missing_count_check,
  c(
    "Missing counts found in the source data.",
    "x" = "Fix missing `count` values before exporting Darwin Core files."
  )
)

stop_if_issues(
  missing_name_check,
  c(
    "Missing species names found in the source data.",
    "x" = "Fix missing `common_name` values before exporting Darwin Core files."
  )
)
```

### Assessment 3: Species name matches the lookup table

```{r qc-species-match}
species_lookup_check <- counts |>
  distinct(common_name) |>
  mutate(
    common_name_key = stringr::str_to_lower(stringr::str_squish(common_name))
  ) |>
  left_join(species_lookup, by = "common_name_key") |>
  filter(is.na(scientific_name))

show_check(species_lookup_check)

stop_if_issues(
  species_lookup_check,
  c(
    "Species names without a match in the lookup table were found.",
    "x" = "Fix the species lookup mismatch before exporting Darwin Core files."
  )
)
```


### Assessment 4: Taxon IDs match the avilistr reference list

```{r qc-avilistr-taxon-id}
taxon_id_check <- counts |>
  distinct(common_name_key) |>
  left_join(species_lookup_join, by = "common_name_key") |>
  distinct(vernacular_name_source, taxon_id) |>
  filter(!is.na(taxon_id)) |>
  anti_join(avilist_lookup, by = "taxon_id")

show_check(taxon_id_check)
```

### Assessment 5: Duplicate taxon records within one event

```{r qc-duplicate-observations}
duplicate_observation_check <- counts |>
  count(date, site, common_name, name = "n_observation") |>
  filter(n_observation > 1)

show_check(duplicate_observation_check)
```

## Records excluded from the export

```{r export-subset}
excluded_missing_count <- counts |>
  filter(is.na(count)) |>
  transmute(site, date, common_name, reason = "Missing count")

excluded_missing_name <- counts |>
  filter(is.na(common_name)) |>
  transmute(site, date, common_name, reason = "Missing common_name")

excluded_duplicate_observation <- counts |>
  add_count(date, site, common_name, name = "n_observation") |>
  filter(n_observation > 1) |>
  transmute(
    site,
    date,
    common_name,
    reason = "Duplicate date + site + common_name"
  )

excluded_export_records <- bind_rows(
  excluded_missing_count,
  excluded_missing_name,
  excluded_duplicate_observation
) |>
  distinct()

show_table(
  excluded_export_records,
  caption = htmltools::tags$caption(
    "Records temporarily excluded from the Darwin Core export."
  )
)
```
