---
title: "Generate GBIF dataset: metadata, figures and tables"
author: "Raphaël Nussbaumer"
#date: `r Sys.Date()`
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(ggplot2)
library(DT)
library(jsonlite)
library(leaflet)
library(mapview)
library(lubridate)
library(RColorBrewer)
library(rgdal)
library(gridExtra)
library(knitr)
library(kableExtra)
```

## Introduction

This script produces the information for the metadata (figures, tables and key numbers) and generate the `events.csv` and `occurances.csv` files in Darwin format for the publication of the [waterbird count data of Sabaki and Mida](https://github.com/A-Rocha-Kenya/Waterbird-counts-Sabaki-Mida) on GBIF (doi).


### Load data

The data are entered and stored in a excel spreadsheet alongside all other waterbird counts performed by [A Rocha Kenya](https://www.arocha.or.ke/) since 1998. Here, we filter for the counts performed at Mida and Sabaki.

At Sabaki, some counts were performed in two teams on each side of the river (north-south). Here, we sums these counts. 

We also load `species.list` and `locations.list` from other sheets which contains more information on species and locations than the main `counts` table. 

```{r}
path = "C:/Users/rnussba1/OneDrive/ARK/ARK - Science/04. Water Bird Count/"

countst <- read_xlsx(paste0(path,"Data/water_bird_count_data.xlsx"), sheet = "Main", col_types = c("text", "numeric","text", "text", "date", "date", "date", "text","text","text","text","text","text","text","text")) %>%
  filter( site=='Sabaki' | site=='Mida Creek' | site=='Sabaki (South)' | site=='Sabaki (North)') %>% 
  replace_na(list(quality="", coverage="", method="",water="",weather="",disturbed="", tidal="",participants="",comment=""))

counts <- countst %>% 
  filter(date<"2004-02-02" & date>"2004-02-01") %>% # WHY? 
  filter(date<"2021-01-01") %>% # Limit dataset publish to the year 2020
  filter(site=='Sabaki (South)' | site=='Sabaki (North)') %>% 
  group_by(date) %>% 
  mutate(start_time = min(start_time),
            end_time = max(end_time)) %>% 
  group_by(common_name,date) %>%
  summarise(count = sum(count),
            quality= first(quality), 
            site = 'Sabaki', 
            start_time = min(start_time),
            end_time = max(end_time),
            coverage = first(coverage),
            method = first(method),
            water = first(water),
            tidal = first(tidal),
            weather = first(weather),
            disturbed = first(disturbed),
            participants = first(participants),
            comment = first(comment),
            .groups = 'drop') %>% 
  full_join(countst %>% filter(site=='Sabaki'| site=='Mida Creek' ),
            by = c("date", "common_name", "count", "quality", "site", "start_time", "end_time", "coverage", "method", "water", "tidal", "weather", "disturbed", "participants", "comment"))

species.list <- read_xlsx(paste0(path,"Data/water_bird_count_data.xlsx"), sheet = "Species")

locations.list <- read_xlsx(paste0(path,"Data/water_bird_count_data.xlsx"), sheet = "Sites") %>% 
  filter( site_name=='Sabaki' | site_name=='Mida Creek')

participants.list <- read_xlsx(paste0(path,"Data/water_bird_count_data.xlsx"), sheet = "Participants")

```

### Basic statistic

We provide a basic description of the dataset with the following table

```{r, echo=FALSE}
counts %>% group_by(site) %>% 
  summarise(
   nbSurvey = n_distinct(date),
   nbSpe = n_distinct(common_name),
   nbObs = n(),
   nbInd = sum(count)
  )  %>% 
  kable() %>% kable_styling() 
  # t() %>% kable("markdown")
```

## Quality control


### Assessement 2: Unique value per event
Check that there is an unique survey (defined as date+site), meaning that all metadata are the same for all the observations of each surveys. 
```{r}
counts %>% 
  group_by(date, site) %>% 
  summarise(disctinct_starttime = n_distinct(start_time),
            disctinct_endtime = n_distinct(end_time),
            disctinct_coverage = n_distinct(coverage),
            disctinct_method = n_distinct(method),
            disctinct_water = n_distinct(water),
            disctinct_tidal = n_distinct(tidal),
            disctinct_weather = n_distinct(weather),
            disctinct_disturbed = n_distinct(disturbed),
            disctinct_participants = n_distinct(participants),
            disctinct_tidal = n_distinct(tidal)
  , .groups = 'drop') %>%
  filter(rowSums( .> 1) > 2 )
```

### Assessement 3: Species validation correct
Check that the name entered havec their equivalence in the Specie tab of the spreadsheet. Often, it is a problem of case-sensitive. 
```{r}
counts %>% 
  group_by(common_name) %>% 
  summarise(.groups = 'drop') %>% 
  left_join(species.list, by="common_name") %>% 
  filter( is.na(scientific_name) )
```

### Assessement 5: Unique observation

Check that Species are entered only once for each survey. 
```{r}
counts %>% 
  group_by(date, site,common_name) %>% 
  filter(n()>1) %>% 
  arrange(date)
```

### Assessement 4: Manual check of Specie and count

Export pivot table for expert assessment of count
```{r}

```

## Dataset Description

### Geographic coverage
```{r, echo=FALSE, message=FALSE, warning=FALSE}

geojson <- fromJSON(paste0(path,"Waterbird-counts-Sabaki-Mida/data/locations.geojson"))

track = readOGR(paste0(path,"Waterbird-counts-Sabaki-Mida/data/locations.geojson"),require_geomType = "wkbLineString") %>% spTransform( CRS("+init=epsg:4326"))
area = readOGR(paste0(path,"Waterbird-counts-Sabaki-Mida/data/locations.geojson"),require_geomType = "wkbPolygon")%>% spTransform( CRS("+init=epsg:4326"))

m <- leaflet(width = "100%", height = "500px") %>% 
  addTiles() %>% 
  addPolygons(data=area,
              stroke=geojson$features$properties$stroke[1],
              weight=geojson$features$properties$`stroke-width`[1],
              opacity=geojson$features$properties$`stroke-opacity`[1],
              fillColor=geojson$features$properties$fill [1],
              fillOpacity=geojson$features$properties$`fill-opacity`[1],
              group = "Area of bird counted") %>% 
  addPolylines(data=track,
              color=geojson$features$properties$stroke[2],
              weight=geojson$features$properties$`stroke-width`[2],
              opacity=geojson$features$properties$`stroke-opacity`[2],
              fillOpacity=geojson$features$properties$`fill-opacity`[2],
              group = "Path followed during transect")

m

mapshot(m, file = paste0(path,"Waterbird-counts-Sabaki-Mida/data/locations.png"))

```

**Bounding box**: `r area@bbox`

## Taxonomic coverage

```{r}
species.list.recorded <- counts %>% 
  group_by(common_name) %>% 
  summarise(
    nbObs = n(), 
    nbInd = sum(count), .groups = 'drop') %>% 
  left_join(species.list, by="common_name") %>% 
  arrange(desc(nbInd))

species.list.recorded %>% head() %>% kable() %>% kable_styling()
```

```{r}
family.list.recorded <- species.list.recorded %>% 
  group_by(family) %>% 
  summarise(nbObs=n(), 
            family_english=first(family_english),
            nbSpecies = sum(nbObs),
            nbInd = sum(nbInd), .groups = 'drop') %>% 
  arrange(desc(nbInd))

family.list.recorded %>% head() %>% kable() %>% kable_styling()

# gsub("\"","",family.list.recorded %>% 
#  transmute(a=paste0('*',family,'* (',family_english,')')) %>% toString())

# family.list.recorded$family_english

```
The dataset includes `r species.list.recorded %>% filter(taxon_rank=="Species") %>% nrow()` unique species and `r species.list.recorded %>% filter(taxon_rank!="Species") %>% nrow()` other taxons (subspecies, family, slash etc.) belonging to `r nrow(family.list.recorded)` families. 



## Temporal coverage


**Temporal range**: `r min(counts$date)` - `r max(counts$date)`

```{r}
for (s in c('Sabaki','Mida Creek')){
dt <- counts  %>% 
  filter( site==s) %>%  
  group_by(date) %>% 
  mutate(month=month(date), year = year(date)) %>% 
  group_by(month, year) %>%
  summarise(nb = n(), .groups="drop")

p1<-ggplot(dt, aes(x = year, y = month, color= nb)) +
  geom_point(size = 7, shape=15) +
  scale_colour_gradientn(colours = brewer.pal(9, 'YlGnBu')) +
  scale_y_continuous(breaks=1:12) +
  scale_x_continuous(breaks=min(dt$year):max(dt$year)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p3<-ggplot(dt, aes(x=year)) + geom_histogram(bins=length(min(dt$year):max(dt$year))) + 
  scale_x_continuous(breaks=min(dt$year):max(dt$year)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p2<-ggplot(dt, aes(x=month)) + geom_histogram(bins=12) + coord_flip() + 
  scale_x_continuous(breaks=1:12)

png(paste0(path,"Waterbird-counts-Sabaki-Mida/data/coverage_",s,".png"), width = 800, height = 400)
grid.arrange(p1,p2,p3,layout_matrix = cbind(c(2,2,2,6), c(1,1,1,3), c(1,1,1,3), c(1,1,1,3)))
dev.off()

grid.arrange(p1,p2,p3,layout_matrix = cbind(c(2,2,2,6), c(1,1,1,3), c(1,1,1,3), c(1,1,1,3)))
}

  
```
## Content providers
```{r}
counts$participants %>% 
  toString() %>% 
  strsplit(",") %>% unlist() %>% trimws() %>% 
  unique() %>% 
  lapply(function(x) participants.list$Name[participants.list$Initial == x]) %>% 
  unlist() %>% sort() %>% 
  toString()
```

## Export in Darwin format


### Export Event table
```{r}

events <- counts %>% 
  group_by(date, site ) %>% 
  summarise(start_time = first(start_time),
            end_time = first(end_time),
            coverage = first(coverage),
            method = first(method),
            water = first(water),
            tidal = first(tidal),
            weather = first(weather),
            disturbed = first(disturbed),
            participants = first(participants), 
            .groups="drop",
  ) %>%
  left_join(locations.list, by=c("site" = "site_name")) %>% 
  transmute(
    type = "Event",
    language = "en",
    license = "http://creativecommons.org/publicdomain/zero/1.0/legalcode",
    rightsHolder = "A Rocha Kenya",
    ownerInstitutionCode = "ARK",
    eventID = paste(sep="_", format(date,"%Y%m%d"), if_else(site=='Sabaki', "sabaki", "mida")),
    samplingProtocol = 'Water Bird Count',
    sampleSizeValue = difftime(end_time, start_time, units = "min"),
    sampleSizeUnit = "minutes",
    samplingEffort = coverage,
    # sampleSizeValue = if_else(site=="Sabaki",3.40,6.06),
    # sampleSizeUnit = "square kilometre",
    eventDate = format(date,"%Y-%m-%d"),
    eventTime = paste0(format(start_time,"%H:%M"),"/",format(end_time,"%H:%M")),
    # eventRemarks = "",
    locationID = if_else(site=='Sabaki', "sabaki", "mida"),
    continent = "Africa",
    # waterBody = if_else(site=='Sabaki', "http://www.geonames.org/197853/galana.html", "https://www.geonames.org/186759/mida-creek.html"),
    country = "Kenya",
    countryCode = "KE",
    # stateProvince =" ",
    county = "Kilifi",
    # municipality = if_else(site=='Sabaki', "Sabaki", "Mida"),
    locality = if_else(site=='Sabaki', "Sabaki River Mouth", "Mida Creek"),
    #locationRemarks = description,
    decimalLatitude = latitude,
    decimalLongitude = longitude,
    # geodeticDatum ="WGS84",
    # footprintWKT = if_else(site=='Sabaki',"POLYGON ((40.1294906 -3.1562848,40.1303489 -3.1575917,40.1319797 -3.1595199,40.1333959 -3.1615125,40.1339323 -3.1633336,40.1359064 -3.1646619,40.1394469 -3.1666116,40.1413138 -3.1699325,40.1402194 -3.1725249,40.1405198 -3.1751601,40.1439531 -3.1781381,40.1459272 -3.1762956,40.1484377 -3.1710251,40.1524717 -3.1611482,40.1489527 -3.1582987,40.1466567 -3.158063,40.1454121 -3.1573882,40.1424188 -3.1545065,40.1388891 -3.1541208,40.1349408 -3.1522033,40.1327843 -3.1529264,40.1335032 -3.1564669,40.1344044 -3.1583844,40.133825 -3.1588772,40.132838 -3.1575274,40.1310784 -3.1552563,40.1294906 -3.1562848))", "POLYGON ((39.9635702 -3.3292204,39.9630981 -3.3312768,39.9632269 -3.333419,39.9617249 -3.3379175,39.9663168 -3.3422017,39.971767 -3.3402738,39.976123 -3.3452649,39.9780434 -3.349517,39.9885738 -3.3435459,39.9846846 -3.3381745,39.9832683 -3.3340188,39.9831825 -3.3328192,39.9836975 -3.3319195,39.9850279 -3.3301201,39.9862724 -3.3288776,39.9878603 -3.3264784,39.9900919 -3.324422,39.9880749 -3.3208231,39.9842983 -3.320866,39.9815517 -3.321123,39.9787622 -3.3212516,39.975844 -3.3218514,39.9724966 -3.3228796,39.9697929 -3.3240364,39.9676901 -3.3256216,39.9654585 -3.3266498,39.9644714 -3.3272068,39.9638277 -3.3283635,39.9635702 -3.3292204))"),
    #georeferencedBy = "Raphaël Nussbaumer",
    #georeferencedDate = "03/06/2020",
    #georeferenceSources = "https://www.geonames.org/ | https://www.google.co.ke/maps/",
    #georeferenceVerificationStatus = "verified by curator",
    #georeferenceRemarks = "",
    dynamicProperties = paste0("{",
      'coverage: "',coverage,'", ',
      'water: "',water,'", ',
      'tidal: "',tidal,'", ',
      'weather: "',weather,'", ',
      'disturbed: "',disturbed,'", ',
      'participants: "',participants,'", ',
      'CWAC_sitecode: "',CWAC_reference,'", ',
      'IBA_sitecode: "',IBA,'", ',
      "}"
    ),
    )

events %>% head() %>%  kable() %>% kable_styling()
```

### Export occurance table
```{r}
occurences <- counts %>% 
  left_join(species.list, by="common_name") %>% 
  transmute(
    basisOfRecord = "HumanObservation",
    eventID = paste(sep="_", format(date,"%Y%m%d"), if_else(site=='Sabaki', "sabaki", "mida")),
    occurrenceID = paste(sep="_",eventID, taxon_id),
    individualCount = count,
    # organismQuantity = count,
    # organismQuantityType = "individu",
    # occurrenceStatus = "present",
    taxonID = taxon_id,
    scientificName = scientific_name,
    kingdom = "Animalia",
    phylum = "Chordata",
    class = "Aves",
    taxonRank = taxon_rank,
    scientificNameAuthorship = "",
    vernacularName = common_name,
    occurrenceRemarks = paste0(quality," | ", comment),
      )

occurences %>% head() %>%  kable() %>% kable_styling()
```

## Write to Excel
```{r}
v <- '11'

write.csv(events, file = paste0(path,"Waterbird-counts-Sabaki-Mida/data/events_",v,".csv"), 
          na = "", row.names = FALSE, fileEncoding = "UTF-8")

write.csv(occurences, file = paste0(path,"Waterbird-counts-Sabaki-Mida/data/occurences_",v,".csv"), 
          na = "", row.names = FALSE, fileEncoding = "UTF-8")
```
